A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-05-11 podc_paper/algorithm-GC.tex
reformatted appendix code

\begin{figure}
\begin{minipage}[t]{0.465\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{200}

@@

\gc{
\Function{Object}{CompleteDeq}{\var{leaf}, $h$}
    \State \linecomment Finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1}$
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\State \Return{\var{res}}
\EndFunction{CompleteDeq}
}

@@

\Function{void}{Append}{\typ{Block} \var{B}} 
    \State \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}}
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}, doing GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $G$}
    	\State \linecomment{Do garbage collection}\label{GCstart}
        \State $s \assign \Call{SplitBlock}{\var{v}}.\var{index}$\label{callSplitBlock}
        \State \Call{Help}{}\label{invokeHelp}
        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < s$}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}

@@

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}

\begin{algorithmic}[1]
\setcounter{ALG@line}{249}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
	\State \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
	\gc{
	\State $T \assign \var{v}.\fld{blocks}$\label{refreshReadTGC}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
		}
    \EndIf

@@

@@

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}
    \State \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State $T\assign \var{v}.\fld{parent.blocks}$
		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa
		\Else
			\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper3}
			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
		\EndIf
	\EndIf
\EndFunction{Propagated}
}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$}
    \State \linecomment{is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated}
    \State \linecomment{to root and $D(\var{v}.\fld{blocks}[\var{b}])$ has at least \var{i} dequeues}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\gc{
		\State $T\assign \var{v}.\fld{parent.blocks}$
		\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper1}
		\State $B_p'$ \assign max block in $T$ with $\edir < b$\label{searchsuper2}
	    }
	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -$ \\
	    		$\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }$\\
					$\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

\end{algorithmic}
\end{minipage}
\vspace*{-4mm}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
\end{figure}

\begin{figure}
\begin{minipage}[t]{0.465\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{302}

\gc{
\Function{void}{Help}{}
    \State \linecomment{help pending operations}

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B} \assign \Call{MaxBlock}{\ell.\fld{blocks}}$
    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ 
        					\State \hspace*{4mm} and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
            \State \linecomment{operation is a propagated dequeue}
            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}
\spac

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} 
    \State\linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State $\var{new}.\fld{index} \assign i$\label{createIndexGC}
    }
	\State \hangbox{\var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}}
	\State \hangbox{\var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} + \\
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \hangbox{\var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + \\
        	\var{num\sub{enq}} - \var{num\sub{deq}}$)}\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmptyGC}
        \State \Return \nl \hfill \com\ no blocks to be propagate to \var{v}\label{CBnullGC}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

\end{algorithmic}
\end{minipage}
\begin{minipage}[t]{0.529\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{351}
\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \hangbox{\var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$\\
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$}\label{FRnumenqGC}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC}
    	\State \linecomment{ queue is empty when dequeue occurs}
        \gc{
        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
        \EndIf
        }
        \State \Return \nl \hfill \label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}
		\State \linecomment{compute enqueue's block using binary search}
		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC}
		\State \linecomment{find rank of enqueue within its block}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$
		\gc{
        \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b_e$
        \EndIf
        \State \Return \var{res}
		}
    \EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}

@@

\begin{algorithmic}[1]
\setcounter{ALG@line}{332}

@@

\caption{Bounded-space queue implementation, continued.\label{pseudocode2GC}}
\end{figure}

>>>>

\begin{figure*}
\begin{minipage}[t]{0.41\textwidth}
\begin{algorithmic}[1]
\setcounter{ALG@line}{200}

@@

\gc{
\Function{Object}{CompleteDeq}{\var{leaf}, $h$} 
    \State \linecomment finish propagated dequeue in $\var{leaf}.\fld{blocks}[h]$
    \State $\langle \var{b}, \var{i}\rangle \assign\ \Call{IndexDequeue}{\var{leaf}, h, 1}$
    \State $\var{res} \assign \Call{FindResponse}{b, i}$
	\State \Return{\var{res}}
\EndFunction{CompleteDeq}
}

@@

\Function{void}{Append}{\typ{Block} \var{B}}  \linecomment{append block to leaf and propagate to root}
    \gc{\State \var{leaf}.\fld{blocks} \assign \Call{AddBlock}{\var{leaf}, \var{leaf}.\fld{blocks}, \var{B}}\label{appendLeafGC}}
    \State \Call{Propagate}{\var{leaf}.\fld{parent}} 
\EndFunction{Append}

@@

\gc{
\Function{RBT}{AddBlock}{\typ{Node} \var{v}, \typ{RBT} \var{T}, \typ{Block} \var{B}} 
    \State \linecomment{add block $\var{B}\neq\nl$ to \var{T}; do GC if necessary}
    \If{\var{B}.\fld{index} is a multiple of $G$}
    	\State \linecomment{Do garbage collection}\label{GCstart}
        \State $s \assign \Call{SplitBlock}{\var{v}}.\var{index}$\label{callSplitBlock}
        \State \Call{Help}{}\label{invokeHelp}
        \State $T' \assign\ \Call{Split}{T,s}$\label{splitGC}
        \State \linecomment{\op{Split} removes blocks with $\fld{index} < s$}
        \State \Return \Call{Insert}{$T',B$}\label{GCend}
	\Else\ \Return \Call{Insert}{$T, B$}
    \EndIf
\EndFunction{AddBlock}
}

@@

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}} \linecomment{try to append a new block to $\var{v}.\fld{blocks}$}
	\gc{
	\State $T \assign \var{v}.\fld{blocks}$\label{refreshReadTGC}
    \State $\var{h} \assign \Call{MaxBlock}{T}.\fld{index}+1$ \label{refreshReadMax}
    }
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlockGC}
    \If{\var{new = \nl}} \Return{\tr} \label{addOPGC}
    \Else  
    	\gc{
	    \State $T' \assign \Call{AddBlock}{v, T, \var{new}}$\label{refreshABGC}
		\State \Return \Call{CAS}{\var{v}.\fld{blocks}, $T$, $T'$}\label{casGC}
		}
    \EndIf

@@

\end{algorithmic}
\end{minipage}
\hfill
\begin{minipage}[t]{0.56\textwidth}

\begin{algorithmic}[1]
\setcounter{ALG@line}{267}

\gc{
\Function{boolean}{Propagated}{\typ{Node} \var{v}, \typ{int} \var{b}}  \linecomment{check if $\var{v}.\fld{blocks}[b]$ has  propagated to \var{root}}
    \State \linecomment{Precondition:  $\var{v}.\fld{blocks}[b]$ exists}
    \If {$\var{v} = \var{root}$} \Return{true}
    \Else
    	\State $T\assign \var{v}.\fld{parent.blocks}$
		\State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
		\If{$\Call{MaxBlock}{T}.\edir < b$} \Return \fa
		\Else
			\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper3}
			\State \Return \Call{Propagated}{\var{v}.\fld{parent}, $B_p$.\fld{index}}
		\EndIf
	\EndIf
\EndFunction{Propagated}
}

@@

\Function{$\langle\typ{int}, \typ{int}\rangle$}{IndexDequeue}{\typ{Node} \var{v}, \typ{int} \var{b}, \typ{int} \var{i}}
    \State \linecomment{return $\langle\var{x, y}\rangle$ such that \var{i}th dequeue in $D(\var{v}.\fld{blocks}[\var{b}])$ is \var{y}th dequeue of $D(\var{root}.\fld{blocks}[\var{x}])$}
    \State \linecomment{Precondition: \var{v}.\fld{blocks}[\var{b}] exists and has propagated to root and $|D(\var{v}.\fld{blocks}[\var{b}])|\geq \var{i}$}
    \If{$\var{v} = \var{root}$} \Return $\langle\var{b, i}\rangle$ \label{indexBaseCaseGC}
    \Else
	    \State \fld{dir} \assign\ (\var{v}.\fld{parent.left} = \var{v} ? \fld{left} : \fld{right}) 
    	\gc{
		\State $T\assign \var{v}.\fld{parent.blocks}$
		\State $B_p$ \assign min block in $T$ with $\edir \geq b$\label{searchsuper1}
		\State $B_p'$ \assign max block in $T$ with $\edir < b$\label{searchsuper2}
	    }
	    \State \linecomment{compute index \var{i} of dequeue in superblock $B_p$}
	    \State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\var{b}-1].\fld{sum\sub{deq}} -
	    		\var{v}.\fld{blocks}[\gc{B_p'}.\edir].\fld{sum\sub{deq}}$}
        \If{$\fld{dir} = \fld{right}$} 
        	\State \hangbox{\var{i} += $\var{v}.\fld{blocks}[\gc{B_p}.\eleft].\fld{sum\sub{deq}} - \mbox{ }
					\var{v}.\fld{blocks}[\gc{B_p'}.\eleft].\fld{sum\sub{deq}}$}\label{considerLeftBeforeRightGC}
        \EndIf \label{computeISuperEndGC}
        \State \Return\Call{IndexDequeue}{\var{v}.\fld{parent}, $\gc{B_p.\fld{index}}$, \var{i}}
    \EndIf
\EndFunction{IndexDequeue}

@@

\gc{
\Function{void}{Help}{} \linecomment{help pending operations}

    \For{$\ell$ in $\var{leaf}[1..k]$}
        \State $\var{B} \assign \Call{MaxBlock}{\ell.\fld{blocks}}$
    	%\State \Call{Propagate}{$\ell.\fld{parent}$}   <--- don't do this because recursive helping could be infinite loop
        \If{$\var{B}.\fld{element} = \nl$ and $\var{B}.\fld{index}>0$ and \Call{Propagated}{$\ell, \var{B}.\fld{index}$}} 
            \State \linecomment{operation is a propagated dequeue}
            \State $\var{B}.\fld{response} \assign\ \Call{CompleteDeq}{\ell, \var{B}.\fld{index}}$ 
        \EndIf
    \EndFor
\EndFunction{Help}
}
\spac

\Function{Block}{CreateBlock}{\typ{Node} \var{v}, \typ{int} \var{i}} \linecomment{create new block  to install in \var{v}.\fld{blocks}[\var{i}]}
    \State let \var{new} be a new \typ{Block} \label{initNewBlockGC}
    \gc{
    \State \var{new}.\eleft \assign\ $\Call{MaxBlock}{\var{v}.\fld{left}.\fld{blocks}}.\fld{index}$\label{createEndLeftGC}
    \State \var{new}.\eright \assign\ $\Call{MaxBlock}{\var{v}.\fld{right}.\fld{blocks}}.\fld{index}$\label{createEndRightGC}
    \State $\var{new}.\fld{index} \assign i$\label{createIndexGC}
    }
	\State \var{new}.\fld{sum\sub{enq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + 
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{enq}}
	\State \var{new}.\fld{sum\sub{deq}} \assign\ \var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{deq}} +
			\var{v}.\fld{right.blocks}[\var{new}.\eright].\fld{sum\sub{deq}}
    \State \var{num\sub{enq}} \assign\ $\var{new}.\fld{sum\sub{enq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{enq}}$\label{computeNumEnqGC}
    \State \var{num\sub{deq}} \assign\ $\var{new}.\fld{sum\sub{deq}} - \var{v}.\fld{blocks}[\var{i}-1].\fld{sum\sub{deq}}$
    \If{$\var{v} = \var{root}$}
        \State \var{new}.\fld{size} \assign\ max(0, $\var{v}.\fld{blocks}[\var{i}-1].\size\ + 
        	\var{num\sub{enq}} - \var{num\sub{deq}}$)\label{computeLengthGC}
    \EndIf
    \If{$\var{num\sub{enq}} + \var{num\sub{deq}} = 0$}\label{testEmptyGC}
        \State \Return \nl \hfill \com\ no blocks to be propagate to \var{v}\label{CBnullGC}
    \Else
        \State \Return \var{new}
    \EndIf
\EndFunction{CreateBlock}

@@

\Function{element}{FindResponse}{\typ{int} \var{b}, \typ{int} \var{i}} \linecomment{find response to \var{i}th dequeue in $D(\var{root}.\fld{blocks}[\var{b}])$}
    \State \linecomment{Precondition:  $1\leq i\leq |D(\var{root}.\fld{blocks}[\var{b}])|$}
    \State \var{num\sub{enq}} \assign\ $\var{root}.\fld{blocks}[\var{b}].\fld{sum\sub{enq}} - \mbox{ }$
    		$\var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}}$\label{FRnumenqGC}
    \If{$\var{root}.\fld{blocks}[\var{b}-1].\size + \var{num\sub{enq}} < \var{i}$}\label{checkEmptyGC} \linecomment{queue is empty when dequeue occurs}
        \gc{
        \If{$b > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b$
        \EndIf
        }
        \State \Return \nl \hfill \label{returnNullGC}
    \Else \ \linecomment{response is the \var{e}th enqueue in the root}
        \State \var{e} \assign\ \var{i} + \var{root}.\fld{blocks}[\var{b}-1].\fld{sum\sub{enq}} - 
			\var{root}.\fld{blocks}[\var{b}-1].\size\label{computeEGC}

		\State find min $b_e \leq \var{b}$ with $\var{root}.\fld{blocks}[b_e].\fld{sum\sub{enq}} \geq \var{e}$\label{FRsearchGC} \linecomment{use binary search}
		\State $i_e \assign\ \var{e} - \var{root}.\fld{blocks}[b_e-1].\fld{sum\sub{enq}}$ \linecomment{find rank of enqueue within its block}
		\gc{
        \State \var{res} \assign \Call{GetEnqueue}{\var{root}, $b_e$, $i_e$}\label{findAnswerGC}
        \If {$b_e > \var{last}[\var{id}]$} $\var{last}[\var{id}]\assign b_e$
        \EndIf
        \State \Return \var{res}
		}
    \EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{minipage}
\vspace*{-4mm}
\caption{Bounded-space queue implementation.\label{pseudocode1GC}  $G$ is a constant, which we choose to be $p^2 \ceil{\log p}$.}
\end{figure*}

\begin{figure*}
\begin{algorithmic}[1]
\setcounter{ALG@line}{341}

@@

\caption{Bounded-space queue implementation, continued.\label{pseudocode2GC}}
\end{figure*}
